Subset coercions in \Coq

### 2006-08-22 papers/article/main.tex
Sorting of sub fix
Christine Fixes, small

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}

@@

@@

\def\sidecond{\sideconden}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a pratical functionnal programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgments can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. We
propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. This solution also provides % PVS:Language,
facilities to express properties with a more mathematical flavor,
bridging a gap between mathematical convention and type theory.

@@

Instead of trying to find a general method for synthetising proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system has been used with great success and
fits naturally with how we write specifications. The main contribution
of this paper is to show how the \ps{} method can be adapted in a proof
assistant with proof terms, which formally justifies the extension in
the first place.

The rest of the article is organized as follows: first we present a
type system based on \CIC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of traductions between
judgements of the two systems. Then, we present a prototype
implementation with some case studies, and finally we give our
conclusions on this work and discuss future directions.

\section{\Russell{}}
The name of our language is an homage to the mathematician Bertrand Russell who
discovered the famous paradox of set theory with the unrestricted
comprehension axiom. In this theory, it is possible to construct the set
$X = \{x `| x `; x\}$ whose definition is circular. Clearly, if $X `: X$
then $X `; X$ and if $X `; X$ then $X `: X$, hence we have an
inconsistency. \Russell{} was one of the pioneers of type theory when he
devised a set theory with a restricted comprehension axiom that
permitted to create subsets only from already defined subsets,
forbidding the definition of $X$. % \cite{whitehead.russell:principia}.

@@

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, you usually prove \emph{first}
that you apply it only to elements having the desired property. 
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. We would like for example
to prove that our usage of partial functions is correct only after the
program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and the
properties which we use to reason about them.

@@

The first one formalizes the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type checking condition} $`G \vdash P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria is rather vague in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalize on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

\subsection{Typing and conversion}
Our presentation of \ps{} differs slightly from the original one
essentially because of the proof terms, but we have also extended it in some ways.

\subsubsection{Type system}
\begin{figure}[t]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}

@@

    \vspace{\infvspace}
    \Subsum\DP

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}
%
\Russell{}'s type system (figure \vref{fig:typing-decl-rules}) is based
on the Calculus of (Co-)Inductive Constructions
\cite{coquandhuet88}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. We omit (co-)inductive
constructs here, as they can be considered as constants and leave their
treatment as future work. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is equivalent to $U$ in environment $`G$. 
Following the presentation as a Pure Type System, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$. As usual, we
let $s, s_i$ for $i `: `N$ range over sorts. The
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
set of rules $\PTSrules$ is defined by the following (functional) relation:
% \begin{table}[h]
%   \vspace{-1.5em}
  \axiomsd
%   \vspace{-1em}
%   \caption{Definition of $\PTSrules$}
%   \label{fig:PTSrules}
% \end{table}

We allow to form dependent sums $\Sigma x : U.V$ (\irule{Sigma}, \irule{Sum-Dep},
\irule{Pi-1}, \irule{Pi-2} rules) only when $U, V : \Set$
or $U, V : \Prop$. The first sum represents dependent pairs, useful
when defining functions returning a tuple, the latter represent
dependent pairs of propositions, which is rarely used. The dependent
pair with $U : \Set$ and $V : \Prop$ is the subset type which we
distinguish (rule \irule{Subset}). The last possible pair $U : \Prop, V : \Set$ is forbidden
because it corresponds to a pair where the last component, which is
informative, may depend on a particular proof of the proposition
$U$. This is quite contrary to the mantra ``Computations do not depend
on proofs'' which governs our programming language.

Likewise, we need to restrict the system so that it becomes property irrelevant for
the propositions appearing in subset types. Indeed, in \PVS{} you can't
manipulate proof terms, so you have
no way to create a term of type $0 \neq 0$ from the judgment $0 :
\mysubset{x}{`N}{P}$, but in a classical Calculus of Constructions with
sigma types or inductive types, the subset type is a first class object
defined with proper introduction and
elimination terms (figure \vref{fig:cci-subset}). What we need to do is forbid the use of these terms
in \Russell{} terms, leaving only the type former.

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \vspace{\infvspace}

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset}
  \vspace{-2.5em}
\end{figure}

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:coerce-impl-rules-short}).

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
(figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
$\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \vref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity rule which is not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
  \TODO{example?}
\end{proof}
\end{paragraph}

Classicaly, we have the following correspondence between the two
systems:
\begin{theorem}[Soundness]
  If $`G \typea t : T$ then $`G \typed t : T$.
  If $`G \typea T \suba{} U : s$ then $`G \typed T \subd U : s$.
\end{theorem}

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then $`G \typea t : T' \suba T$.
  If $`G \typed T \subd U : s$ then $`G \typea T \suba U : s$.
\end{theorem}

Finaly, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}

@@

We can now prove the following:
\begin{theorem}[Forgetful map correctness]
  If $`G \typec t : T$ then $`G^{\circ} \typea t^{\circ} : T^{\circ}$ if $()^{\circ}$ is
  defined (ie, does not return $`_$) on $`G, t$ and $T$.
\end{theorem}

Practically, this means we can use almost all existing definitions in the
\Coq environment. Obviously, this map is involutive and
the identity on \Russell terms. 
%, potentially turning them (locally) to an opaque form
% if their body is not traductible
\section{From \lng{} to \CICq{}}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CIC
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

@@

Now we can build a total interpretation:
\begin{proposition}[Interpretation correctness]
  If $`G \typea t : T$ then $\ipG{`G} \typeq \ip{t}{`G} : \ip{T}{`G}$.
\end{proposition}

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organized as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the precedent
algorithm. Finally we will present a proof sketch of the aforementionned
proposition.

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which build a coercion given two types. 
The side-conditions of the rules have not changed, so we omit them for
improved readability.

@@

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instanciation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied. Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and well suited for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respect some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

We suppose that conversion is preserved by the interpretation $\ip{}{}$, so
the rule \irule{SubConv} derives an empty coercion, as the target system
will be able to derive itself that $\ip{T}{`G} `= \ip{U}{`G}$.
Similarly for \irule{SubHnf}, we rely on the target's system conversion
rule so that the coercion $c$ of codomain $\ip{\hnf{U}}{`G}$ can be seen
as an object of type $\ip{U}{`G}$.

@@

\interpfig[Interpretion of terms]

@@

\paragraph{Aknowledgments}
I'd like to thank Christine Paulin-Mohring for directing this work.

\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}

\renewcommand{\thefootnote}{}
\footnotetext{This article was typeset in \LaTeX~using the
  \texttt{Concrete Math} font}
\end{document}

>>>>

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}

@@

\newenvironment{code}{\begin{alltt}\vspace{-1em}}{\end{alltt}\vspace{-1em}}

@@

\def\id#1{\texttt{#1}}
\def\sidecond{\sideconden}

@@

\abstract{We propose a new language for writing programs with dependent
  types which can be elaborated into partial \Coq terms. This language permits
  to establish a phase distinction between writing and proving
  algorithms in the \Coq{} environment.  Concretely, this means allowing
  to write algorithms as easily as in a pratical functionnal programming
  language whilst giving them as rich a specification as desired and
  proving that the code meets the specification using the whole \Coq{}
  proof apparatus. This is achieved by extending conversion to an
  equivalence which relates types and subsets based on them, a technique
  originating from the ``\emph{Predicate subtyping}'' feature of \PVS{}
  and following mathematical convention. The typing judgments can be
  translated to the \CICfull (\CIC{}) by means of an interpretation
  which inserts coercions at the appropriate places. These coercions can
  contain existential variables representing the propositional parts of
  the final term. A prototype implementation of this process has been
  implemented.}

\section{Introduction}
There are many means to program in the \Coq{} environment. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent complex types, we immediatly get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{head} which returns
the first element of a list. In \ML{}:
\begin{code}
  let head = function hd :: tl -> hd | [] -> assert(false)
\end{code}

This function is partial, its domain is reduced to non-empty lists. In
\Coq, we would rather write the following:
\begin{code}
Definition head (l : list A) : option A := 
  match l with
    hd :: tl -> Some hd
  | [] -> None
  end.
\end{code}

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, using a
richer specification: 
\begin{code}
Definition head (l : list A) : forall p : l <> [] -> A.
\end{code} 
Now the definition's body requires some heavy plumbing of the code which is 
not afordable when programming. We propose a solution to overcome this difficulty, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure along with an elaboration into partial \Coq terms. Using
\Russell, we would write:
\begin{code}
  Definition head \{ l : list A | l <> [] \} : A := 
  match l with
    hd :: tl -> hd
  | [] -> _
  end
\end{code}

The specification shows that we're definining a partial function yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proof in the second branch. After type-checking,
the proof assistant will ask the user to prove that in the context where
\id{l} is a non-empty list and $l = []$, we can create an object of type
$A$ (which should be obvious as the context is contradictory).

This solution also provides % PVS:Language,
facilities to express properties with a more mathematical flavor using subsets,
bridging a gap between mathematical convention and type theory.

@@

Instead of trying to find a general method for synthetising proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system has been used with great success and
fits naturally with how we write specifications using the subset type \cite{progmartinlof}.
The main contribution of this paper is to show how the \ps{} method 
can be adapted in a proof assistant with proof terms, which formally 
justifies the extension in the first place.

The remainder of the article is organized as follows: in section \ref{sec:Russell} we present a
type system based on \CIC, which integrates subset ``subtyping'', and
we show how it relates to \CIC{} by means of traductions between
judgements of the two systems in section \ref{sec:Traduction}. Then, we present a prototype
implementation with some case studies, and finally we give our
conclusions on this work and discuss future directions.

\section{\Russell{}}
\label{sec:Russell}
The name of our language is an homage to the mathematician Bertrand Russell who
discovered the famous paradox of set theory with the unrestricted
comprehension axiom. In this theory, it is possible to construct the set
$X = \{x `| x `; x\}$ whose definition is circular. Clearly, if $X `: X$
then $X `; X$ and if $X `; X$ then $X `: X$, hence we have an
inconsistency. \Russell{} was one of the pioneers of type theory when he
devised a set theory with a restricted comprehension axiom that
permitted to create subsets only from already defined subsets,
forbidding the definition of $X$. % \cite{whitehead.russell:principia}.

@@

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, e.g: $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and the
properties which we use to reason about them.

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset}
  \vspace{-2.5em}
\end{figure}

@@

The first one formalizes the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type checking condition} $`G \vdash P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria can be rather large in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalize on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

\subsection{Typing and conversion}
% Our presentation of \ps{} differs slightly from the original one
% essentially because of the proof terms (hence the use of a judgement
% instead of a function for checking subtyping), but we have also extended
% it by . 
\subsubsection{Type system}
\begin{figure}[t]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}

@@

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-2em}
  \caption{\CIC typing judgement}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

\begin{figure}[t]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}
    \vspace{\infvspace}
    \Subsum\DP

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell new rules}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgement $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $(s_1, s_2, s_2)$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ _ : unit `| A \}, B$ products to benefit from
our languages subtyping mechanism.

We allow to form dependent sums $\Sigma x : U.V$ (\irule{Sigma},
\irule{Sum-Dep}, \irule{Pi-1}, \irule{Pi-2} rules) only when $U, V :
\Set$ or $U, V : \Prop$. The first sum represents dependent pairs,
useful when defining functions returning a tuple, the latter represent
dependent pairs of propositions, which is most often used as conjunction
of propositions. The dependent pair with $U : \Set$ and $V : \Prop$ is
the subset type which we distinguish (rule \irule{Subset}). The last
possible pair $U : \Prop, V : \Set$ is forbidden because it corresponds
to a pair where the last component, which is informative, may depend on
a particular proof of the proposition $U$. This is quite contrary to the
mantra ``Computations do not depend on proofs'' which governs our
programming language.

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgment $0 : \mysubset{x}{`N}{P}$, but
in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). What
we need to do is forbid the use of these terms in \Russell{} terms,
leaving only the type former (figure \vref{fig:typing-decl-rules}).

The reductions rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ it's head normal form.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. 
% This notion is presented in \PVS{} by
% means of a $\mu_0$ operator which was readily adaptable in our setting
% (figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
% new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
% $\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
\end{proof}
\end{paragraph}

As usual, we have the following correspondence between the two
systems:
\begin{theorem}[Soundness]
  If $`G \typea t : T$ then $`G \typed t : T$.
  If $`G \typea T \suba{} U : s$ then $`G \typed T \subd U : s$.
\end{theorem}

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then $`G \typea t : T'$ and $`G \typea T' \suba T : s$.
  If $`G \typed T \subd U : s$ then $`G \typea T \suba U : s$.
\end{theorem}

Finaly, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
\begin{proof}
  Rules are syntax-directed.
\end{proof}

We denote by $\typeafn{`G}{t}$ the function which returns the type of
term $t$ in context $`G$.

@@

We can now prove the following:
\begin{theorem}[Forgetful map correctness]
  If $`G \typec t : T$ then $`G^{\circ} \typea t^{\circ} : T^{\circ}$ if $()^{\circ}$ is
  defined (ie, does not return terms containing $`_$) on $`G, t$ and $T$.
\end{theorem}

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms. Indeed, the second projection of a
subset element can only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it is forbidden.

\section{From \lng{} to \CICq{}}
\label{sec:Traduction}
We will now build an interpretation
$\ip{t}{`G}$ of \Russell terms $t$ in environment $`G$ into \CIC
terms. We will check at the end that it respects the $()^{\circ}$ operator
in the sense that if $t$ is well-typed in $`G$ then $\ip{t}{`G}^{\circ} = t$.

@@

We are ready to build the interpretation, which should have the following property:
\begin{proposition}[Interpretation correctness]
  If $`G \typea t : T$ then $\ipG{`G} \typeq \ip{t}{`G} : \ip{T}{`G}$.
\end{proposition}

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organized as follows: first we will
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we will define the
interpretation of terms which is mutually recursive with the precedent
algorithm. Finally we will present a proof sketch of the aforementioned
proposition.

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which build a coercion given two types (we denote it $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
improved readability. The judgement $\subimpl{`G}{c}{T}{U}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{`G}{T}$ to an object of type $\ip{`G}{U}$
in context $\ipG{`G}$. We use the algorithmic types to drive the
derivation but the resulting object will be a well-typed $\Coq$ term. We
could also directly use the interpreted types but that complicates the
proof.

@@

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instanciation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied. Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and well suited for checking coherence 
of coercions, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

As conversion will be preserved by the interpretation $\ip{}{}$,
the rule \irule{SubConv} derives an empty coercion, as the target system
will be able to derive itself that $\ip{T}{`G} `= \ip{U}{`G}$.
Similarly for \irule{SubHnf}, we rely on the target's system conversion
rule so that the coercion $c$ of codomain $\ip{\hnf{U}}{`G}$ can be seen
as an object of type $\ip{U}{`G}$.

@@

\interpfigs[Interpretion of terms]

@@

\paragraph{Aknowledgments}
I'd like to thank Christine Paulin-Mohring for directing this work.
\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}
% \renewcommand{\thefootnote}{}
% \footnotetext{This article was typeset in \LaTeX~using the
%   \texttt{Concrete Math} font}

\end{document}
